>

### 2017-01-04 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
More work on article draft

---
title: >
    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing to incorporate output from ensembles of different climate models and different simulations of each model. This processing can be particularly extensive when identifying and characterizing extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps and may use identification threholds based on local distributions (e.g., 98\textsuperscript{th} percentile temperature for that location). Here, we provide an overview of working with daily climate model output data in R. Further, we present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. This package takes as input a directory of projection files from one or many climate models, identifies all occurrences of a specific type of extreme event (e.g., heat waves) for multiple cities using a user-specified event definition, and writes an output directory of files with these extreme events and their characteristics (e.g., length, absolute and relative temperatures). The \pkg{futureheatwaves} package then allows the user to create custom functions and apply them across all event files; this functionality can be used either to explore extreme event patterns (e.g., average length and intensity across different climate models) or to apply epidemiological estimates to project health impacts under different scenarios.
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

@@

Research on climate change impacts can require extensive processing. This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1}, in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments. Projections of regional climate change over the next century are subject to high uncertainty due to three distinct sources: (1) internal climate variability, i.e. climate noise, (2) climate model uncertainty, i.e. the same forcing can produce a different response in different models and (3) scenario uncertainty, i.e. uncertainty in future climate forcings (e.g. \citet{hawkins2009potential}). One approach for ensuring that these sources of uncertainty are characterized is to simulate the future climate many times with multiple models and for multiple future scenarios.

For example, the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}) brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

Processing climate output data is particularly intensive for data in smaller time steps, like daily output. While some climate impacts can be assessed using climate model output at monthly, seasonal, or yearly time steps, the impacts of multi-day extreme events must be assessed using output in daily time steps. Such multi-day extreme events include heat waves, cold spells, frost day spells, and droughts. The assessment is further complicated if extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}, as this event definition must be determined at each study location from climate model output. Further, it is often of interest to create summaries of multiple characteristics of these extreme events. For example, one may be interested in determining whether the frequency or characteristics (e.g., length, intensity) of heat waves or warm spells will change under certain climate change scenarios \citep{IPCCch1}.

We begin this article with an overview of climate model output data, particularly daily data, for R users. We outline where data from CMIP5 can be obtained as well as how to work with the file format (netCDF) from R. We overview some R packages that can be useful when working with this data, as well as aspects of the data (e.g., non-standard calendars) of which users should be aware when working with daily climate model output in R.

After this overview, we present the \pkg{futureheatwaves} package which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}). This package can be used to identify and characterize a variety of multi-day extreme events, but provides some functionality particularly useful in identifying and characterizing heat waves specifically. Quantification of the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature. These include: (1) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario) and (2) uncertainty in the definition of a heat wave itself. Thus, identifying and characterizing the impacts of future heat waves in state-of-the-art climate models requires analyzing hundreds of projections based on combinations of anthropogenic activity, climate model and ensemble member, adaptation scenario, and heat wave definition. Such an analysis is non-trivial. For example, to estimate the possible impacts of heat waves using 32 ensemble members of CMIP5, two forcing scenarios, two different heat wave definitions, and five progressive assumptions about adaptation to heat, would require that one identify and characterize heat waves in 640 separate temperature time series for each region of interest.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large sets of climate projections more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models, and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate projection, to allow the exploration of patterns in these characteristics across different projections and also to allow the use of more complex impacts models, including models that include effect modification by event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate projections by applying custom functions across heat wave data sets from all projections at once.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

The \pkg{futureheatwaves} package automates the process of identifying and characterizing multi-day extreme events across different ensemble members of one or more climate models. A variety of different heat wave definitions have been used to identify heat waves in a time series of temperature data \citep{smith2013heat}, and the choice of heat wave definitions can influence both health effect estimates \citep{chen2015influence, kent2014heat} and projected heat wave trends \citep{smith2013heat}. Further, other types of extreme events will be defined differently than heat waves (for example, frost day spells may be defined as one or more days with temperature at or below 32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output. Finally, this package allows the user to explore the extreme events identified in the climate model output with a function that can take a user-defined R function and apply it across all extreme event files generated for the separate ensemble members.

@@

For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5). Over 20 climate modeling groups have created one or more climate models which, for this project, are run using standardized scenarios \citep{taylor2012overview}. The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}. For CMIP5, each group ran simulations under several experiments, with experiments varying in terms of the radiative forcing. This radiative forcing depends on time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.), which are specified for each experiment \citep{taylor2012overview, IPCCch9}. Experiments include historical experiments (run using forcings consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}. Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions, resulting in multiple ensemble members for a single climate model and experiment. CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature). Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).

The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, this output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Filenames for CMIP5 files can be parsed to generate information about the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}. Files can be searched and downloaded through a point-and-click web interface. The data portal also allows you to generate a specific wget script, which you can use to download many files at once. A wget script can also be generated through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF). NetCDF is a binary file format that allows storage of data representing a regular array. These climate model output files can be as large as several gigabytes [?]. For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude) (Figure \ref{fig:netcdfexample}). Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}. All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}. This required metadata includes the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time).

\begin{widefigure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{widefigure}

\citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources for finding out more about the CMIP climate model output data.

@@

A few R packages can be used to work with the netCDF file format used for CMIP5 files. Earlier packages to work with netCDF files included \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. Climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5} and so should work with any of these packages, although it is safer to write code using functions that can be used with either version in case future phases of CMIP do not require files to conform with netCDF version 3.

The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory. The metadata describes the dimensions of the array and the units, etc., of each variable and is printed by the print method of the R object returned by the \code{nc\_open} function of \pkg{ncdf4}. For CMIP5 files, the dimension variables include the times and locations corresponding to each array in the data. Before reading in a variable from a netCDF file, a connection to the file must be opened, for example with the \code{nc\_open} function from \pkg{ncdf4}. Variables can then be read in using the \code{ncvar\_get} function from \pkg{ncdf4}, with the \code{varid} parameter set to "lat", "lon", or "time" (as a caveat, many climate models output to non-Gregorian calendars, in which case the time variable should be read in using a different function, as discussed later in this section). The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. You can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years. Some climate models, however, are run using different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}. With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values. The \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers} packages provide further functionality with netCDF files that can be useful when working with climate model output data and provide particular help in working with different calendars. CMIP5 netCDF files include information on the calendar used, which the \code{nc.get.time.series} function in \pkg{ncdf4.helpers} use to convert the "time" variable in the file to an object of the PCICt class, which provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}. As a note, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the PCICt object correctly matches the expected date ranges for a file as you process CMIP5 data in R.

The size of the climate model output files can be large enough that it may make more sense to work with smaller chunks of the data, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}. This problem aggregates when working with multiple climate models and more than one ensemble member for each of those climate models.

The functions and packages described can be used to work with CMIP5 netCDF files in R to do things like map near-surface air temperatures from a single climate model on a specific day in July 2075 (Figure \ref{fig:worldmap}) and pull a time series of daily near-surface air temperature simulations for the climate model grid point closest to Beijing, China (Figure \ref{fig:timeseries}). The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.

@@

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member.}
\label{fig:timeseries}
\end{figure}

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special "cmip5data" class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data (e.g., the functions \code{checkTimePeriod} and \code{cmip5data}, as well as functions that work with \code{cmip5data} objects including \code{filterDimensions} and \code{getProjectionMatrix}) \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load the daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe that parses the information contained in the file name. As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function. However, this function does not allow downloading of climate model output with finer time steps, like daily data. This package uses the \code{models2wux} function to read in climate model output netCDF files and convert to a WUX dataframe that can be used by other functions in the package. While this function can input climate model output with daily time steps, if the element "what.timesteps" of the \code{modelinput} list input is set to "daily", the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

More tips on working with CMIP5 files in R are provided in the "starting_from_netcdf" vignette of the \pkg{futureheatwaves} package.

>>>>

---
title: >
    Working with daily climate model output data in R and the \pkg{futureheatwaves} package
author:
  - name: G. Brooke Anderson
    affiliation: Colorado State University
    address:
    - Department of Environmental & Radiological Health Sciences
    - 1681 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  brooke.anderson@colostate.edu
  - name: Colin Eason
    affiliation: Colorado State University
    address:
    - Department of Computer Science
    - 1873 Campus Delivery
    - Fort Collins, Colorado 80523
    email:  aimesce@gmail.com
  - name: Elizabeth A. Barnes
    affiliation: Colorado State University
    address:
    - Department of Atmospheric Science
    - 1371 Campus Delivery
    - Fort Collins, CO 80523
    email:  eabarnes@atmos.colostate.edu
abstract: >
  Research on climate change impacts can require extensive processing, especially when using ensemble techniques to incorporate output from different climate models and different simulations of each model. This processing can be particularly extensive when identifying and characterizing multi-day extreme events like heat waves and frost day spells, as these must be processed from model output with daily time steps. Further, climate model output is in a format and follows standards that may be unfamiliar to most R users. Here, we provide an overview of working with daily climate model output data in R. We then present the \pkg{futureheatwaves} package, which we developed to ease the process of identifying, characterizing, and exploring multi-day extreme events in climate model output. 
preamble: >
  % Supposedly allows you to add stuff to LaTeX preamble. Not for me.
output: rticles::rjournal_article
---

@@

Research on climate change impacts can require extensive processing of climate model output data. This is not only because output files for a single climate model can be large, but also because of the rising popularity of ensemble techniques \citep{IPCCch1} in which, to better characterize uncertainty in projections, impacts are assessed for multiple climate models, multiple simulations of each climate model, and multiple climate experiments. Such ensemble techniques help characterize uncertainty in projections of regional climate change over the next century due to three distinct sources: (1) internal climate variability, i.e. climate noise, (2) climate model uncertainty, i.e. the same forcing can produce a different response in different models and (3) scenario uncertainty, i.e. uncertainty in future climate forcings (e.g. \citet{hawkins2009potential}).

A key source of data for ensemble techniques is the Coupled Model Intercomparison Project, phase 5 (CMIP5; \citet{taylor2012overview}). This project brought together dozens of major climate modeling groups around the world to simulate the same future radiative forcing scenarios, but with their own models. This created an ensemble of state-of-the-art climate model projections that allow researchers to study projections and their uncertainties. Most of these modeling groups additionally performed more than one simulation for each scenario and model (i.e. multiple ensemble members), perturbing the initial conditions by a very tiny amount to quantify uncertainties due to internal climate variability.

We begin this article with an overview of CMIP5 climate model output data for R users, focusing on output with a daily time step. We outline where data from CMIP5 can be obtained as well as how to work with the file format (netCDF) from R. We overview some R packages that can be useful when working with this data, as well as aspects of the data (e.g., non-standard calendars) of which users should be aware when working with daily climate model output in R.

After this overview, we present the \pkg{futureheatwaves} package, which we created to aid in identifying and characterizing any type of multi-day extreme event from daily climate model output (Table \ref{tab:goals}). The impacts of multi-day extreme events must be assessed using output in daily time step, unlike other climate impacts that can be assessed using climate model output at monthly, seasonal, or yearly time steps. Further, extreme events are identified based on conditions that are rare for a certain location (e.g., 98th percentile of local temperature distribution for identifying heat waves) \citep{IPCCch1}. In this case, assessment is further complicated, as the event definition must be determined at each study location from climate model output. Finally, it is often of interest to create summaries of multiple characteristics of these extreme events. For example, one may be interested in determining whether the frequency or characteristics (e.g., length, intensity) of heat waves or warm spells will change under certain climate change scenarios \citep{IPCCch1}.

The \pkg{futureheatwaves} package handles these challenges and can be used to identify and characterize a variety of multi-day extreme events across different ensemble members of one or more climate models. It also provides some functionality particularly useful in identifying and characterizing heat waves specifically. Quantification of the impacts of heat waves on human health suffer from additional sources of uncertainty beyond those inherent in projections of regional changes in surface temperature. These include: (1) uncertainty in the definition of a heat wave itself and (2) uncertainty in the ability of communities to adapt to changing temperatures (the adaptation scenario). This package therefore allows the user to create and use a custom extreme event definition to identify events in the climate model output, as well as options to explore different scenarios of adaptation to heat.

\begin{table}[t]
\begin{center}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\begin{tabular}{lL{\dimexpr0.8\textwidth-2\tabcolsep\relax}}
\toprule
 & Design goals of the \pkg{futureheatwaves} package \\
\midrule
1 & Make processing of large ensembles of climate simulations more practical for researchers exploring the potential impacts of heat waves and other multi-day extreme events.\\
2 & Speed up processing time by incorporating C++ in event identification.\\
3 & Keep track of the names of climate models and number of ensemble members processed for each.\\
4 & Not only identify, but also characterize, all extreme events within each climate simulation, to allow the exploration of patterns in these characteristics across different simulations and also to allow the use of more complex impacts models, including models that incorporate event characteristics (e.g., event length, event intensity). For example, this package allows the user to apply a health effects model where risk of mortality is not the same for every heat wave, but rather is modified by heat wave length, intensity, or other measured characteristics.\\
5 & Give users extensive power in customizing the process, including allowing custom extreme event definitions.\\
6 & Allow users to easily explore the extreme events identified within all climate simulations.\\
7 & Create output that is in a ``tidy" data format, allowing it to work well with \pkg{ggplot2} for visualization.\\
\bottomrule
\end{tabular}
\end{center}
\caption{Design goals for the \pkg{futureheatwaves} package.}
\label{tab:goals}
\end{table}

@@

For climate impact studies, a main source for climate model output is the Coupled Model Intercomparison Project, which is currently in its fifth phase (CMIP5). Over 20 climate modeling groups created one or more climate models which, for this project, were run using standardized scenarios \citep{taylor2012overview}. The resulting output is uniform across modeling groups and has a consistent structure, which allows comparison of simulations from different models \citep{IPCCch9}. CMIP5 climate model output is archived at a number of different time steps (e.g., daily, monthly, seasonal, yearly) \citep{taylor2010cmip5}, and some variables are reported at multiple levels in the ocean or atmosphere (e.g., ocean temperature). Here, we will focus on data with a daily time step for variables reported at a single level (e.g., near-surface air temperature).

Each modeling group ran simulations under several experiments, with experiments varying in terms of radiative forcing by using different scenarios of time-varying model inputs (greenhouse gas emissions or concentrations, land use changes, etc.) \citep{taylor2012overview, IPCCch9}. Experiments include historical experiments (run using radiative forcing consistent with observed and reconstructed data for 1850--2005), pre-Industrial control experiments, and experiments of future scenarios of radiative forcing over the 21st century or longer (e.g., RCP4.5, RCP8.5) \citep{taylor2012overview}. Some modeling groups created ensembles of output for a specific model and experiment, in which they ran the experiment multiple times with the model with very small changes to the initial conditions.

The CMIP5 climate model output data is distributed across data nodes at different climate modeling centers \citep{taylor2012overview}, but can be accessed centrally at the World Climate Research Programme CMIP5 data portal at https://pcmdi.llnl.gov/search/cmip5/. Users must register before downloading data, and some data are restricted to non-commercial use. There is a separate file for each combination of climate model, experiment, modeling realm (e.g., atmosphere, ocean), variable, time step, and ensemble member \citep{taylor2012overview, taylor2010cmip5}. For finer time scales, the output is further split across multiple files for specific year ranges (e.g., 5 years of output for each file) \citep{taylor2010cmip5}. Each file's name includes the output variable, climate model, experiment, and ensemble member for the simulation \citep{taylor2010cmip5}.

CMIP5 files can be searched and downloaded through a point-and-click web interface. They can also be downloaded in bulk to computers with Unix or Mac operating systems using the \code{wget} file downloading utility. Appropriate \code{wget} scripts can be created either through the World Climate Research Programme CMIP5 data portal or through Earth System Grid Federation's Search RESTful API. Tips on efficiently searching and downloading the data, including through use of wget scripts and the search API, are available as user tutorials through the website of the University of Colorado Boulder's Earth System CoG (e.g., https://www.earthsystemcog.org/projects/cog/doc/wget for a tutorial on downloading files using wget).

CMIP5 files are saved in Network Common Data Format (netCDF), a binary file format that allows storage of data representing a regular array. For climate model output at a single level (e.g., near-surface air temperature), the data is a 3-dimensional array, with dimensions representing time and two coordinates of location (e.g., latitude and longitude). Figure \ref{fig:netcdfexample} provides a sketch of the structure of netCDF files for single-level climate model output. Global climate models generate output at regularly-spaced time steps, typically at regularly-spaced grid points around the world. The latitude and longitude spacing of grid points vary by climate model, but are typically 1--2 degrees for atmospheric variables in CMIP5 models \citep{IPCCch9}. Each data point in the netCDF array gives the modeled value of the variable (e.g., surface temperature) for a single time point and location. For CMIP5 climate model output, the location units are in degrees east and degrees north for longitude and latitude, respectively. For daily output files, the time unit is in days since a specified origin date-time (e.g., days since 1850-01-01 00:00:00) \citep{taylor2010cmip5}. All CMIP5 output files are required to include certain metadata  \citep{taylor2010cmip5}, including the experiment, forcing agents input to the model to create the simulation, time step, institution and institutional contact information, climate model, and modeling realm \citep{taylor2010cmip5}. The metadata also must include units for all of the coordinate variables (e.g., longitude, latitude, time). The netCDF format allows you to access metadata and variables describing the dimensions of the data without reading the full file into memory.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{netcdf_figure}
\end{center}
\caption{Example of structure of a NetCDF climate model output file for a variable reported at a single depth, like surface air temperature. Data are stored in a three-dimensional array, with measurements at each time step and grid location. Surface temperature data are typically indexed in climate model output files by longitude, latitude, and time, in that order. For example, if the air surface temperature ("tas") is read into an R object called \code{tas}, you can access the value for the first day at the fourth longitude and third latitude with \code{tas[4, 3, 1]}. In addition to the output variable (temperature in this example), vectors with the ordered values of each dimension (longitude, latitude, and time) can also be read in from the netCDF file, as well as attribute data (e.g., units for variables, the calendar used for time).}
\label{fig:netcdfexample}
\end{figure}

To find out more about the CMIP climate model output data, \citet{taylor2012overview} and \citet{meehl2007wcrp} are excellent resources.

@@

When working with daily climate model output data, challenges to R users include: (1) the file format; (2) use of non-Gregorian calendars; and (3) large file sizes. This section explains these challenges and offers some strategies for dealing with them.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{worldmapexample}
\end{center}
\caption{Example of mapping near-surface air temperature data worldwide for a single day of climate model output data. This map uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member, on a single day in the summer of 2075. Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
\label{fig:worldmap}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width = 0.8\textwidth]{timeseriesexample}
\end{center}
\caption{Example of plotting a time series of temperature simulations between 2071 and 2075 from CMIP5 daily climate model output data for the model grid cell point closest to Beijing, China. This plot uses data from the Geophysical Fluid Dynamics Laboratory's Earth System Model 2G, r1i1p1 ensemble member. Full code for recreating the map is available in the "starting\_from\_netcdf" vignette of the \pkg{futureheatwaves} package.}
\label{fig:timeseries}
\end{figure}

CMIP5 data are available as netCDF files, for which base R import functions do not exist. There are, however, a few R packages that can be used to work with the netCDF file format used for CMIP5 files. Older packages include \pkg{ncdf} and \pkg{ncvar}, but these do not work with the newer netCDF version 4 released in 2008 and are no longer available through CRAN. More recent packages, including \pkg{ncdf4} \citep{ncdf4} and \pkg{RNetCDF} \citep{michna2013rnetcdf, RNetCDF}, work with both version 4 and netCDF's older version 3. While climate model output data for CMIP5 is required to conform with the earlier version (version 3) \citep{taylor2010cmip5}, it is safer to write code using functions that can be used with either version, in case future phases of CMIP do not require files to conform with netCDF version 3.

You can do a number of things with netCDF files in R using these functions. For example, \pkg{ncdf4}'s \code{nc\_open} function can be used to open a connection to a netCDF file, and the object returned by the function includes the file's attribute data. Once a file connection is open, variables can be read in using the \code{ncvar\_get} function. For example, the variables defining the dimensions of the sketched netCDF file in Figure \ref{fig:netcdfexample} could be read into R with \code{ncvar\_get} with the \code{varid} parameter set to "lat", "lon", or "time". The climate output variable (e.g., near-surface air temperature) can similarly be read in using \code{ncvar\_get}. In this case, the \code{varid} parameter should be set using the appropriate CMIP5 variable name (e.g., "tas" for near-surface air temperature); these variable names can be found in the CMIP Requested Output tables \citep{taylor2010cmip5}. If you only need data from a subset of the full file, you can use the dimensional time and location data to identify the location of the variable data you need in the netCDF array and use indexing to read only that data into memory, without needing to read in the full file (for example, with the \code{nc.get.var.subset.by.axes} function in \pkg{ncdf4.helpers}). Once the user is done reading in data from the file, the connection can be closed (e.g., with the \code{nc\_close} function from \pkg{ncdf4}).

A second challenge when working with climate model output data in R is that some climate models output to non-Gregorian calendars. Since the late 1500s, Western dates have been set using the Gregorian calendar, which has 365.2425-day years. Some climate models, however, are run using different calendars, including the Julian calendar (365.25-day years), a calendar where there are no leap years (365-day years), a calendar where every year is a leap year (366-day years), and a calendar of twelve 30-day months (360-day years) \citep{cfconventions}. With these non-Gregorian calendars, R's base functions for converting a vector to a Date class based on the number of days since an origin date (\code{as.Date}, \code{as.POSIXct}) do not return the desired values.

Two R packages provide help with non-Gregorian calendars: \pkg{PCICt} \citep{PCICt} and \pkg{ncdf4.helpers} \citep{ncdf4.helpers}. The \code{nc.get.time.series} function in \pkg{ncdf4.helpers} pulls and uses metadata on the calendar stored in the CMIP5 netCDF file's attributes to convert the "time" variable in the file to an object of the \code{PCICt} class. This class is defined in the \pkg{PCICt} package and provides Date-like functionality for 360- and 365-day calendars \citep{PCICt}. However, while these functions will help with handling most CMIP5 files, the CMIP5 standards allows use of other calendars which may not be succesfully handled by these functions, so it is important to assess whether the time variable range in the \code{PCICt} object correctly matches the expected date ranges for a file when processing CMIP5 data in R.

Finally, the size of CMIP5 files can make them difficult to work with in R. CMIP5 climate model output files can be as large as several gigabytes. The size of the files can therefore be large enough that it may make more sense to work with smaller chunks of the data in R, rather than reading all data into memory and working with the data all at once \citep{RCMIP5}. This problem aggregates when working with multiple climate models and more than one ensemble member for each of those climate models.

The \pkg{RCMIP5} package includes functions to read in CMIP5 data from netCDF files, scan a directory of CMIP5 files and determine models with continuous available data, create objects of a special \code{cmip5data} class to work with CMIP5 data within R, and parse the filenames for all files in a directory to extract information within the filename. For this package, most functions only work with monthly or less frequent data \citep{RCMIP5}. While the \code{loadCMIP5} function does successfully load daily data as a \code{cmip5data} object, most of the methods for this object type do not do anything meaningful for daily data. The package's \code{getFileInfo} function, however, will work with CMIP5 files of any time step; this function identifies all CMIP5 files in a directory and creates a dataframe with information parsed from the file name. As a note, the \code{get.split.filename.cmip5} function in the \pkg{ncdf4.helper} package similarly can be used to parse information contained in CMIP5 file names \citep{ncdf4.helpers}.

The \pkg{wux} package \citep{wux} includes functions that allow the user to download CMIP5 monthly-aggregated output directly from within R with the \code{CMIP5fromESGF} function. The package then uses the \code{models2wux} function to read in climate model output netCDF files and convert them to "WUX" dataframes, which can be used by other functions in the package. While this function can input climate model output with daily time steps (the "what.timesteps" element of the \code{modelinput} list input must be set to "daily"), the function aggregates this data to a monthly or less frequent (e.g., seasonal) aggregation when creating the WUX dataframe. Therefore, while this package provides very useful functionality for working with averaged output of daily climate model output data, it cannot easily be used to identify and characterize multi-day extreme events like heat waves.

The functions and packages described in this section can be used with CMIP5 netCDF files to do things in R like map near-surface air temperatures from a single climate model on a specific day (Figure \ref{fig:worldmap}) or pull a time series of daily near-surface air temperature simulations at a specific climate model grid point (Figure \ref{fig:timeseries}). The "starting_from_netcdf" vignette that comes with the "futureheatwaves" package provides all code required to create these figures, as well as more details and code examples on working with CMIP5 netCDF files in R.

@@
